Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2020-12-14 jss/EpiCompare-paper/EpiCompare-paper.tex

\documentclass[
  shortnames]{jss}

@@

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}

@@

The recent (and currently on-going) COVID-19 global epidemic has
galvanized public interest in understanding more about infectious
disease modeling highlighted the usefulness of research in the area of
infectious disease epidemiology. Infectious disease models typically
attempt to do one or more of the following: 1) predict the spread of
current and future epidemics
\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g. ...][]{}. The COVID-19 epidemic highlights how all three
goals are important both separately and taken as a whole. Infectious
diseases inflict enormous burdens on the world: millions of lives lost
and trillions of dollars spent yearly. Correctly analyzing and
addressing these issues aids in prevention and mitigation of future
outbreaks.

The current epidemic of COVID-19 also highlights that these models are
just one piece of the overall analysis pipeline. University based
resources like John Hopkin's and government numerical dashboards (across
all levels of government) during the COVID-19 epidemic remind us that
descriptive statistics and visualization can be a important first step
in the process (multiple \cite{}?).\} Still, rightly so, a large amount
of theoretical work goes into modeling epidemics, with different models
focusing at the individual / agent level, network structure or just
aggregate flows (review paper \cite{}?). All placing individuals /
proportions of the populations into different states (e.g.~suspectible,
exposed, infected, recovered, etc.). With all these models, review and
comparison papers in the literature and through MIDAS (Models of
Infectious Disease Agent Study) Control Center helps the individual
practitioner decide the correct approach.

@@

This package also emphasizes the value of approach epidemics in a
\textit{time-invariant} way, not constrained to things like a specific
time scale, or initial date of the first infection. Although epidemics
are by definition a process that evolves over time, epidemics, including
the recent COVID-19, often need to be compared in a time-invariant way
to understand the processes at play. Additionally, many tools to examine
the quantity of the population in each state along the infection process
(for example: quantity of suspectible vs infected vs recovered
individuals) don't always as intelligently capture the natural
connections between the proportion of individuals in these states. Tools
in \pkg{EpiCompare} attempt to give the user the ability to extend their
toolkit to evaluate epidemics to also include time-invariant approaches.

@@

SHANNON: I FEEL LIKE WE CAN JUST CLEAN UP TO ORIGINAL SECTION AND PUT IT
BACK IN HERE

@@

\begin{CodeChunk}
\begin{CodeInput}
R> with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))
\end{CodeInput}
\begin{CodeOutput}
[1] 0
\end{CodeOutput}
\end{CodeChunk}

The real power of \code{agents_to_aggregate()} lies in its ability to
aggregate over any number of pre-specified states. For example, the
Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the
time of infection and recovery, respectively of each individual. We can
then plot the SIR values through a time-invariant lens using
\pkg{ggplot2} functions or with our custom \code{geom},
\code{geom_aggregate}, which takes the original agent data as input
(Fig. \ref{fig:hag-tern-raw}).

@@

Along with multiple epidemic states, the function
\code{agents_to_aggregate} can also be extended to populations with
vital dynamics (e.g.~birth and death) and examples of this are shown in
the package vignette. In summary, \code{agents_to_aggregate()} is a
multi-purpose workhorse that be leveraged to convert individual level
records into aggregate information that may be more useful for some
forms of epidemic modeling such as compartment modeling.

@@

We first try to model the Hagelloch data with an SIR model (see Eq.
\ref{eq:sir}). In our vignette, we show how to fit a stochastic model
via maximum likelihood and simulate from the model with those best fit
parameters. Our function \code{simulate_agents()} generates individual
level data according to discrete multinomial draws, which depend on the
number of individuals in each state at the previous time step and a
matrix of transition probabilities. For example

@@

The result of our simulation is the object \code{agents} which is a
18800 \(\times\) 5 which details the time of entry into the \(S\),
\(I\), and \(R\) states for a given simulation. Before we examine the
results of this simple SIR model, we will also examine another, more
sophisticated SIR model, this time from the package \pkg{EpiModel}.
Briefly, this model fits a contact network, where the class of the
student is a covariate and then simulates a SIR-epidemic on that
network.

>>>>

% To remove when JSS.cls is fixed
% https://github.com/rstudio/rticles/issues/329
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[
  shortnames]{jss}

@@

% Pandoc citation processing

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{xcolor}

@@

The recent (and currently on-going) COVID-19 global pandemic has
galvanized public interest in understanding more about infectious
disease modeling and has highlighted the usefulness of research in the
area of infectious disease epidemiology. Infectious disease models
typically attempt to do one or more of the following: 1) predict the
spread of current and future epidemics
\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g. ...][]{}. The COVID-19 pandemic highlights how all three
goals are important both separately and taken as a whole. Infectious
diseases inflict enormous burdens on the world: millions of lives lost
and trillions of dollars spent yearly. Correctly analyzing and
addressing these issues aids in prevention and mitigation of future
outbreaks. \textcolor{orange}{really like this paragraph}

The current epidemic of COVID-19 also highlights that infectious disease
models are only one piece of the overall analysis pipeline. University
based resources like John Hopkin's and government numerical dashboards
(across all levels of government) during the COVID-19 epidemic remind us
that descriptive statistics and visualization can be a important first
step in the process (multiple \cite{}?).\} Still, rightly so, a large
amount of theoretical work goes into modeling epidemics, with different
models focusing at the individual / agent level, network structure or
just aggregate flows (review paper \cite{}?). All placing individuals /
proportions of the populations into different states (e.g.~suspectible,
exposed, infected, recovered, etc.). With all these models, review and
comparison papers in the literature and through MIDAS (Models of
Infectious Disease Agent Study) Control Center helps the individual
practitioner decide the correct approach.
\textcolor{orange}{along with expertise from healthcare professionals...}

@@

This package also emphasizes the value of approach epidemics in a
\textit{time-invariant} way, not constrained to things like a specific
time scale, or initial date of the first infection. Although epidemics
are by definition a process that evolves over time, epidemics, including
the recent COVID-19, often need to be compared in a time-invariant way
to understand the processes at play. Additionally, many tools to examine
the quantity of the population in each state along the infection process
(for example: quantity of suspectible vs infected vs recovered
individuals) don't always as intelligently capture the natural
connections between the proportion of individuals in these states. Tools
in \pkg{EpiCompare} attempt to give the user the ability to extend their
toolkit to evaluate epidemics to also include time-invariant approaches.
The goal of \pkg{EpiCompare} is not to supplant existing infectious
disease modeling tools and software but, rather, is a concerted effort
to create standard and fair comparisons among models developed for
disease outbreaks and outbreak data.

@@

\hypertarget{visualize}{%
\subsection{Visualize}\label{visualize}}

One issue that the COVID-19 outbreak has brought to the forefront is the
importance of figures and graphs to inform, support, and even persuade
lay people of technical and scientific results \cite{}. Quantities and
trends shown in these visualizations include exponential growth curves,
S-curves, peak day and percent of infectious indviduals, ICU capacity,
and deaths over time and uncertainty of such estimates.

\pkg{EpiCompare} supports traditional visualizations (e.g.~\# in
condition vs.~time) and also introduces a novel visualization of
epidemics via time free analysis. In particular, we use ternary plots
that show Susceptible-Infectious-Recovered data/models (SIR). The
advantages of these ternary plots are 1) a standard way to compare
epidemics with different outbreak times and time scales, 2) a novel
interpretation of \(R_0\), the initial reproduction number under some
circumstances, and 3) clear visualization of joint SIR uncertainty via
conformal bands.

Ternary plots are sometimes used in chemistry \citep{} but only rarely
seen in the field of infectious disease epidemiology (to our knowledge,
only seen in \cite{}). Ternary plots illustrate the relationship of
constrained 3D data, namely of points
\((a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]\) with
\(a_i+ b_i + c_i =1\) and as such, are situational. The SIR framework,
composed of data and statistical models where individuals pass from
Susceptible to Infectious to Recovered states is a common, if not the
most common, modeling framework in infectious disease epidemiology (seen
in examples of recent published works in \cite{}), and also happens to
be of the situational format required for ternary plots where
\((S(t), I(t), R(t))\) are the number of susceptible, infectious, and
recovered individuals in each state at time \(t\) and
\(S(t) + I(t) + R(t) = N(t)\) where \(N(t)\) is the population size at
time \(t\).

Ternary plots are time-free in the sense that the temporal scale does
not explicitly apear in the visualization and as a consequence, allow
for direct comparisons of outbreaks on different time scales (e.g.~days
vs.~years) or of SIR data of the same disease in different areas (our
first case study in Sec. \ref{sec:ex1} focuses on this).

Ternary plots also, under certain conditions, have a special
relationship with \(R_0\), the initial reproduction number, which has
been called the ``most important quantity in epidemiology'\,' \citep{}.
The reproduction is itself a time-free concept and is the average number
of secondary infections primary infection will generate when the primary
infection is added to a completely susceptible population. To see the
importance of \(R_0\), one need only read the newspaper \citep{} or look
at the length of the table of estimated \(R_0\) quantities for COVID-19
\citep{}.

In regards to traditional \(X\) vs.~time plots, \(R_0\) is difficult to
visualize. For example, as seen in Fig. \ref{fig:r0-trad}, it is
possible that two models have the same value of \(R_0\) but there is no
way of knowing that from looking at the graph.

It is possible to identify, at a glance, which epidemic paths have the
same value of \(R_0\) via ternary plot visualizations, under some
circumstances. To show, this we first need to introduce the classic,
deterministic Kermack and McKendrick SIR model where the initial number
in each state \((S(0), I(0), R(0))\) are known and \(N\) the population
size is constant. The movement of individuals among the states is given
by the following differential equations (Eq. \eqref{eq:sir-ode}) where
\(\beta\) is the average infection rate and \(\gamma\) is the average
recovery rate, \begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align}

If we have two SIR models that follow the Kermack and McKendrick SIR
model, have the same percentage of individuals in the initial states,
and the same value of \(R_0\) then the two models will create exactly
overlapping paths in a ternary plot.

More formally, let two Kermack and McKendrick (see Eq.
\eqref{eq:sir-ode}) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\)
and \((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume
both models have initial values \((S(0), I(0), R(0))\). Let
\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). Equivalently,
\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
some \(a > 0\).

\begin{theorem}\label{thm:sir-scale}
Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{} and is shown in detail in Appendix \ref{app:proof}.
The consequence of Theorem \ref{thm:sir-scale} is that for two SIR
models that have the same initial percent of individuals in each state
and \(R_0\) then for every point on the epidemic path of the first SIR
model is also a point on the epidemic path of the second SIR model.

@@

The real power of \code{agents_to_aggregate()} lies in its ability to
aggregate over any number of pre-specified states. For example, the
Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the
time of infection and recovery, respectively of each individual. We can
then plot the SIR values through a time-invariant lens using
\pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
\ref{fig:hag-tern-raw}) or with our custom \code{geom},
\code{geom_aggregate}, which takes the raw agent data as input.

@@

Along with multiple epidemic states, the function
\code{agents_to_aggregate} can also be extended to populations with
vital dynamics (e.g.~birth and death) and examples of this are shown in
the package vignette. In summary, \code{agents_to_aggregate()} is a
multi-purpose workhorse that may be leveraged to convert individual
level records into aggregate information that may be more useful for
some forms of epidemic modeling such as compartment modeling.

@@

We first try to model the Hagelloch data with an SIR model (see Eq.
\ref{eq:sir}). In our vignette, we show how to fit a stochastic model
via maximum likelihood and simulate from the model with those best fit
parameters. Our function \code{simulate_agents()} generates individual
level data according to discrete multinomial draws, which depend on the
number of individuals in each state at the previous time step and a
matrix of transition probabilities. For example, the below code
generates 100 simulations of an outbreak of a diseease with one initial
infector in a population of \(n= 188\) individuals.

@@

The result of our simulation is the object \code{agents} which is a
18800 \(\times\) 5 which details the time of entry into the \(S\),
\(I\), and \(R\) states for a given simulation. Before we examine the
results of this simple SIR model, we will also examine another, more
sophisticated SIR model, this time from the package \pkg{EpiModel}.
Briefly, this model first fits a contact network to the set of
indivduals, where the class of the student is a covariate. The model
then simulates a SIR-epidemic on that network.
